Best Practices in Latent Class Analysis using Free Open Source Software

### 2022-10-16 manuscript.tex
Update 16 October
Changed title and short title. Adjusted abstract, Added keywords. Made the use of "LCA" and "mixture models" more consistent based on whether we refer to the whole group of LCA techniques, or a mixture subset.  Minor adjustment to wording throughout the paper. Shortened lines throughout. Capitalized titles and made them uniform. Adjusted section-subsection-subsubsection depth (# vs ## vs ###).  Removed irrelevant comments. Wrote best practices in visualization (##Visualization).  Added the tutorial (now set to eval = F for knitting, however code chunks do not show when using

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  ,man]{apa6}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Best Practices in Mixture Modeling using Free Open Source Software},
  pdfauthor={Caspar J. van Lissa1,2},
  pdflang={en-EN},
  pdfkeywords={keywords},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
\usepackage[bidi=basic]{babel}
\else
\usepackage[bidi=default]{babel}
\fi
\babelprovide[main,import]{english}
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{keywords\newline\indent Word count: X}
\DeclareDelayedFloatFlavor{ThreePartTable}{table}
\DeclareDelayedFloatFlavor{lltable}{table}
\DeclareDelayedFloatFlavor*{longtable}{table}
\makeatletter
\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
\makeatother
\usepackage{lineno}

\linenumbers
\usepackage{csquotes}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

\title{Best Practices in Mixture Modeling using Free Open Source Software}
\author{Caspar J. van Lissa\textsuperscript{1,2}}
\date{}

\shorttitle{BEST PRACTICES MIXTURE MODELING}

@@

\abstract{%
Latent class analysis is a popular technique for identifying groups in data based
on a parametric model. Examples of this technique are known as mixture models, latent profile
analysis, latent class analysis, growth mixture modeling, and latent class
growth analysis. Despite the popularity of this technique, there is limited
guidance with respect to best practices in estimating and reporting mixture
models. Moreover, although user-friendly interfaces for advanced mixture
modeling have long been available in commercial software packages, open
source alternatives have remained somewhat inaccessible. This tutorial
describes best practices for the estimation and reporting of latent class analysis,
using free and open source software in R. To this end, this tutorial
introduces new functionality for estimating and reporting mixture mixture
models in the \texttt{tidySEM} R-package. These functions rely on estimation using
the OpenMx R-package.
}

@@

Latent class analysis is an umbrella term that refers to a number of techniques for estimating unobserved group membership
based on a parametric model of one or more observed indicators of group membership.
This method has become quite popular across scientific fields, and under a number of different names; most notably (finite Gaussian) mixture modeling and latent profile analysis.
Despite the popularity of the method, there is a lack of standards for estimating and reporting latent class analyses.
This complicates manuscript review and assessment of the quality of published studies,
and introduces a risk of misapplications of the technique.
The present paper seeks to address this gap in the literature by suggesting updated guidelines for estimation and reporting on latent class analysis, based on current best practice.
Importantly, in order to lower the barrier of entry and ensure reproducibility of all examples, this paper exclusively relies on free, open source software for latent class analysis in R.
Our goal is to make best-practices latent class analysis widely accessable.

\hypertarget{defining-latent-class-analysis}{%
\subsection{Defining latent class analysis}\label{defining-latent-class-analysis}}

Latent class analysis can be understood as a method for estimating unobserved groups
based on a parametric model of observed indicators of group membership.
The concept of latent class analysis can be understood in different ways.
Generally speaking, it can be said that a mixture model assumes that the study population
is composed of \(K\) subpopulations, or classes.
It further assumes that the observed data are a mixture of data generated by class-specific models.
The simplest univariate ``model'' is a normal distribution, which can be described with two parameters:
the mean and the variance.
Commonly, the same model is estimated accross all classes, but with different parameters for each class (i.e., class-specific means and variances).
Mixture modeling then estimates both the parameters for each class, and the probability for each that an individual belongs to each class.

@@

The distribution is evidently bimodal, which bodes well for the intended mixture model.
In this case, the number of classes is known a-priori.
When estimating a two-class mixture model, the detective observes that the model estimates
the mean shoesize of the two groups are equal to 7.25 and 9.22,
which is close to the true means of the two groups, namely 9.04 and 6.93.
When tabulating estimated group membership against observed (known) group membership, it can be seen that women are classified with a high degree of accuracy, but men are not:

@@

One way to is to conceptualize latent class analysis is by analogy to a measurement model in structural equation modeling.
A mixture model is like confirmatory factor analysis, except that the continuous latent variable is substituted with a categorical latent variable.
One difference between the two techniques is that factor analysis can be considered as a way to group observed \emph{variables} into latent constructs, with factor loadings indicating which items belong are most indicative of a construct.
By contrast, mixture modeling groups \emph{individuals} into classes (see Nylund-Gibson \& Choi, 2018).
In line with this distinction, latent class analysis is sometimes referred to as a ``person-centered'' technique, and factor analysis as a ``variable-centered'' technique.

When the focus is on the model parameters in each group,
then latent class analysis can be thought of as similar to a multi-group structural equation model.
The main distinction is that group membership is not known a-priori,
but is instead estimated - with measurement error - based on the data.
Whereas in a multigroup model, the data are split by group and treated as independent samples,
in a mixture model, all cases contribute to the estimation of all parameters in all groups.
The relative contribution of each case to the parameters of each group is determined by that case's posterior probability of belonging to that group.

When the focus is on each individual's estimated class membership,
latent class analysis can be thought of as a type of clustering algorithm.
In line with this perspective, mixture modeling is sometimes described as ``model-based
clustering'' (Hennig, Meila, Murtagh, \& Rocci, 2015; Scrucca, Fop, Murphy, \&
Raftery, 2017).
Many clustering algorithms apply some recursive splitting algorithm to the data.
By contrast ``model-based'' clustering refers to the fact that latent class analysis
estimates cluster membership based on a parametric model.
Specifically, the posterior class probability that an individual belongs to a latent class
can be computed from the likelihood of that individual's observed data under given the class-specific model.

Finally, in the context of machine learning, latent class analysis can be considered as an
\emph{unsupervised classification} problem (Figueiredo \& Jain, 2002).
The term \emph{unsupervised} refers to the fact that the outcome variable, true class membership,
is not known, and the term \emph{classification} refers to the fact that the algorithm is predicting a categorical outcome (class membership).

\hypertarget{a-taxonomy-of-latent-class-analyses}{%
\subsection{A taxonomy of latent class analyses}\label{a-taxonomy-of-latent-class-analyses}}

In this paper, we use the term latent class analysis to refer to techniques that estimate latent class membership based on a parametric model of observed indicators.
From a historical perspective, the term latent class analysis was initially conceived to refer to analyses with categorical (binary) indicators (Vermunt, J.K. et al., 2004).
Nowadays, there are a number of related techniques, known by distinct names, that serve a similar purpose.
The term ``latent class analysis'' seems most appropriate as an umbrella term for this broader class of models,
as it only refers to the purpose of the analysis, and does not imply restrictions to the model used, or the level of measurement of the indicators.
Given the abundance of terms in use for closely related classes of models, we will provide a rudimentary taxonomy of latent class analyses.

One common type of latent class analyses is the \emph{finite Gaussian mixture model};
a univariate analysis where the observed distribution of a single variable is assumed to result from a mixture of a known number of Gaussian (normal) distributions.
The parameters of a finite Gaussian mixture model are the means and variances of these underlying normal distributions.
The analysis of shoe sizes presented earlier is a canonical example of this type of analysis.
In the multivatiate case, with more than one indicator variable,
the parameters of a mixture model are the means, variances, and covariances between the indicators (which can be standardized to obtain correlations).
These parameters can be estimated freely, or constrained, across classes.

The technique known as \emph{latent profile analysis (LPA)} is a special case of the mixture model,
which assumes conditional independence of the indicators.
Conditional independence means that,
after class membership is accounted for,
the covariances/correlations between indicators are assumed to be zero.
This can be conceived of as a restricted mixture model with covariances fixed to zero.
In some cases, such constraints will be inappropriate;
for example, when the cohesion between indicators is expected to differ between classes.
As an example, a mixture model analysis of ocean plastic particles found two classes of particles based on length and width:
A class of smaller particles with a high correlation between length and width, meaning that these particles were approximately round or square in shape, and a class of larger particles with a low correlation between length and width, meaning that these particles were heterogenous in shape.
From a theoretical perspective, this makes sense, because the smaller particles have been ground down to a more uniform shape by the elements.

It is also possible to estimate a mixture model based on latent indicators.
This means that, within each class, one or more continuous latent variables are estimated based on the observed indicators.
Categorical latent variable membership is then estimated based on these continuous latent variables.
A common application of this approach is in longitudinal research, where the indicators reflect one construct assessed at different time points.
Examples of this approach include \emph{growth mixture models} (GMM) and \emph{latent class growth analyses} (LCGA).
These techniques estimate a latent growth model to describe individual trajectories over time.
The growth mixture model is a latent class model where the parameters that indicate class membership
are the intercepts and variances (and typically covariances) of the latent growth variables, e.g., a latent intercept and slope.
This technique assumes that individuals within a class can have heterogenous trajectories.
If the variance of the growth parameters is fixed to zero, it is known as a latent class growth analysis.
This latter approach assumes that all individuals within a class share the same identical trajectory,
and that any variance in the indicators not explained by the class-specific latent trajectories is due to residual error variance.

The term latent class analysis originally referred to cases where the observed indicators were categorical.
Nowadays, it is more commonly used as an umbrella term.
To prevent ambiguity, the special case where indicators are of binary or ordinal measurement level might be described as latent class analysis with ordinal indicators.
Latent class models with ordinal indicators are parameterized differently from mixture models.
One common parameterization assumes that each categorical variable reflects an underlying standard normal distribution.
The parameters are ``thresholds'' that correspond to quantiles of a standard normal distribution (with \(N(\mu = 0, \sigma = 1)\)).
These thresholds are estimated based on the proportion of individuals in each of the response categories of the indicator variable.
For example, a binary indicator has a single threshold that distinguishes the two response categories.
If responses are distributed 50/50, then the corresponding threshold would be \(t_1 = 0.00\).
If the responses are distributed 60/40, then the resulting threshold would be \(t_1 = 0.25\).
This paper will primarily focus on mixture models and special cases thereof,
although most of the suggested guidelines are applicable to latent class analyses.

\hypertarget{use-cases-for-latent-class-analysis}{%
\subsection{Use cases for latent class analysis}\label{use-cases-for-latent-class-analysis}}

There are several use cases for which latent class analyses are suitable.
One example is to test a theory that postulates the existence of a categorical latent variable.
For example, \emph{identity status theory} posits that, at any given point in time, adolescents reside in one of four identity statuses.
Latent class analysis can be used to identify these four statuses based on observed indicators (e.g., self-reported identity exploration and commitment).
If results indicate that the data are better described by a different number of classes,
or that the four-class solution does not correspond to the predicted pattern of responses on the indicators, then the theory may be called into question.

Another use case is unsupervised learning;
when the goal is to restore unobserved class membership based on observed indicators,
or to classify individuals.
For example, a mixture model can be used as a diagnostic aid when several clinical indicators can be used to distinguish between a fixed number of physical (Baughman, Bisgard, Lynn, \& Meade, 2006) or mental (Wu, Woody, Yang, Pan, \& Blazer, 2011) health problems.
The example of shoe size is a rudimentary illustration of this type of application.

\hypertarget{best-practices}{%
\section{Best practices}\label{best-practices}}

\hypertarget{in-estimation}{%
\subsection{In estimation}\label{in-estimation}}

The best practices in estimation, as outlined in Table \ref{tab:checkest},
are rooted in existing recommendations for best practices for estimating
specific sub-types of latent class analyses,
including latent class growth analysis (\textbf{schootGRoLTSChecklistGuidelinesReporting2017?})
and latent class analysis with ordinal indicators (e.g., Nylund-Gibson \& Choi, 2018).
These were generalized to be more relevant to all types of latent class analyses,
and updated to current best practices, as explained below.

\begin{tabular}[t]{l|l}
\hline
\# & Item\\
\hline
1. & Examine observed data\\
\hline
2. & Handling missing data\\
\hline
3. & Alternative model specifications\\
\hline
4. & Starting values\\
\hline
5. & Algorithm\\
\hline
\end{tabular}

\hypertarget{examine-observed-data}{%
\subsubsection{Examine observed data}\label{examine-observed-data}}

Examining observed data is essential for any analysis,
as it may reveal patterns and violations of assumptions that had not been considered prior to data collection.
Pay special attention to level of measurement of the indicators.
Mixture modeling (including LPA) is strictly speaking only suitable for continuous variables.
Indicators with an ordinal level of measurement are likely to violate the assumption of within-class normal distributions of mixture models (see Vermunt, 2011).
Personal experience consulting on latent class analyses and moderating the \texttt{tidyLPA} Google group
suggest that the mis-application of mixture models to ordinal (e.g., Likert-type) indicators is the most common source of user error.
Whereas it has been argued that some parametric methods are robust when scales with 7+ indicators are treated as continuous (e.g., Norman, 2010),
this certainly does not imply that all methods are.
It is certainly unlikely that such ordinal variables can be treated as a \emph{mixture} of multiple normal distributions.
The problem becomes egregious when the number of classes estimated equals or exceeds the number of categories; in this case, each class-specific mean could describe a single response category, and a class-specific variance component would be nonsensical.
In sum, Likert-type scales are rarely suitable for mixture modeling;
latent class analysis with ordinal indicators is more appropriate.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  ,man,floatsintext]{apa6}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
\usepackage[bidi=basic]{babel}
\else
\usepackage[bidi=default]{babel}
\fi
\babelprovide[main,import]{english}
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{latent class analysis, mixture models, best practices, free open source software, tidySEM\newline\indent Word count: 8206}
\usepackage{csquotes}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Best Practices in Latent Class Analysis using Free Open Source Software},
  pdfauthor={Caspar J. van Lissa1,2},
  pdflang={en-EN},
  pdfkeywords={latent class analysis, mixture models, best practices, free open source software, tidySEM},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Best Practices in Latent Class Analysis using Free Open Source Software}
\author{Caspar J. van Lissa\textsuperscript{1,2}}
\date{}

\shorttitle{BEST PRACTICES LCA}

@@

\abstract{%
Latent class analysis is a family of techniques for identifying groups in data based
on a parametric model. Its examples include mixture models, latent profile
analysis, latent class analysis with ordinal indicators, and latent class
growth analysis. Despite the popularity of this technique, there is limited
guidance with respect to best practices in estimating and reporting mixture
models. Moreover, although user-friendly interfaces for advanced mixture
modeling have long been available in commercial software packages, open
source alternatives have remained somewhat inaccessible. This tutorial
describes best practices for the estimation and reporting of latent class analysis,
using free and open source software in R. To this end, this tutorial
introduces new functionality for estimating and reporting mixture
models in the \texttt{tidySEM} R-package whose backend relies on
the \texttt{OpenMx} R-package.
}

@@

Latent class analysis (LCA) is an umbrella term
that refers to a number of techniques for estimating unobserved group membership
based on a parametric model of one or more observed indicators of group membership.
The types of LCA have become quite popular across scientific fields,
most notably finite Gaussian mixture modeling and latent profile analysis.
Vermunt, J.K. et al. (2004) defined LCA more generally as virtually any statistical model
where ``some of the parameters {[}\ldots{]} differ across unobserved subgroups''.

Despite its popularity, there is a lack of standards for estimating and reporting LCA.
While Van De Schoot, Sijbrandij, Winter, Depaoli, and Vermunt (2017) developed reporting guidelines
for a specific type of LCA known as latent growth models,
general reporting guidelines for all LCA are still lacking.
This complicates manuscript review and assessment of the quality of published studies,
and introduces a risk of misapplications of the technique.
The present paper seeks to address this gap in the literature
by suggesting updated guidelines for estimation and reporting of LCA,
based on current best practices.
Importantly, in order to lower the barrier of entry and ensure reproducibility of all examples,
this paper exclusively relies on free, open source software in R.
Our goal is to make best-practices in LCA widely accessible.

\hypertarget{defining-latent-class-analysis}{%
\subsection{Defining Latent Class Analysis}\label{defining-latent-class-analysis}}

Latent class analysis is a group of methods for estimating unobserved groups
based on a parametric model of observed indicators of group membership.
The concept of LCA can be understood in different ways.
A mixture model assumes that the study population
is composed of \(K\) subpopulations or classes.
It further assumes that the observed data are a mixture of data generated by class-specific models.
The simplest univariate model is a normal distribution,
which can be described with two parameters: the mean and the variance.
Commonly, the same model is estimated across all classes,
but with different parameters for each class (i.e., class-specific means and variances).
Mixture modeling then estimates both the parameters for each class,
and the probability that an individual belongs to each class.

@@

The distribution is evidently bimodal, which bodes well for the intended mixture model.
In this case, the number of classes is known a-priori.
When estimating a two-class mixture model, the detective observes that the model estimates
the mean shoe size of the two groups are equal to 7.25
and 9.22,
which is close to the true means of the two groups,
namely 9.04 and 6.93.
When tabulating estimated group membership against observed (known) group membership,
it can be seen that women are classified with a high degree of accuracy, but men are not:

@@

A LCA is like confirmatory factor analysis,
except that the continuous latent variable is substituted with a categorical latent variable.
One difference between the two techniques is that
factor analysis can be considered as a way to group observed \emph{variables} into latent constructs,
and LCA groups \emph{individuals} into classes.
In line with this distinction, LCA is sometimes referred to as a ``person-centered'' technique,
and factor analysis as a ``variable-centered'' technique (Nylund-Gibson \& Choi, 2018).

When the focus is on the model parameters in each group,
LCA can be thought of as similar to a multi-group structural equation model.
The main distinction is that group membership is not known a-priori,
but is instead estimated -- with measurement error -- based on the data.
Whereas in a multi-group model, the data are split by group and treated as independent samples,
in LCA, all cases contribute to the estimation of all parameters in all groups.
The relative contribution of each case to the parameters of each group
is determined by that case's posterior probability of belonging to that group.

When the focus is on each individual's estimated class membership,
LCA can be thought of as a type of clustering algorithm.
In line with this perspective, LCA is sometimes described as ``model-based
clustering'' (Hennig, Meila, Murtagh, \& Rocci, 2015; Scrucca, Fop, Murphy, \& Raftery, 2016).
Many clustering algorithms apply some recursive splitting algorithm to the data.
By contrast ``model-based'' clustering refers to the fact that LCA
estimates cluster membership based on a parametric model.
Specifically, the posterior class probability that an individual belongs to a latent class
can be computed from the likelihood of that individual's observed data under given the class-specific model.

Finally, in the context of machine learning, LCA can be considered as an
\emph{unsupervised classification} problem (Figueiredo \& Jain, 2002).
The term \emph{unsupervised} refers to the fact that the outcome variable
-- true class membership -- is not known,
and the term \emph{classification} refers to the fact that
the algorithm is predicting a categorical outcome -- class membership.

\hypertarget{a-taxonomy-of-latent-class-analysis-methods}{%
\subsection{A Taxonomy of Latent Class Analysis Methods}\label{a-taxonomy-of-latent-class-analysis-methods}}

In this paper, we use the term latent class analysis to refer to
a family of techniques that estimate latent class membership
based on a parametric model of observed indicators.
From a historical perspective, the term was initially conceived to
refer to analyses with categorical, usually binary indicators (Vermunt, J.K. et al., 2004).
Nowadays, there are a number of related techniques,
known by distinct names, that serve a similar purpose.
The term ``latent class analysis'' seems most appropriate
as an umbrella term for this broader class of models,
as it only refers to the purpose of the analysis,
and does not imply restrictions to the model used,
or the level of measurement of the indicators.
Given the abundance of terms in use for closely related classes of models,
we will provide a rudimentary taxonomy of LCA methods.

A common type of LCA is the \emph{finite Gaussian mixture model};
a univariate analysis where the observed distribution of a single variable
is assumed to result from a mixture of a known number of normal (Gaussian) distributions.
The parameters of a finite Gaussian mixture model are
the means and variances of these underlying normal distributions.
The analysis of shoe sizes presented earlier is a canonical example of this type of analysis.
In the multivariate case, with more than one indicator variable,
the parameters of a mixture model are the means, variances, and covariances
between the indicators (which can be standardized to obtain correlations).
These parameters can be estimated freely, or set to be constrained across classes.

The technique known as \emph{latent profile analysis (LPA)} is a special case of the mixture model,
which assumes conditional independence of the indicators.
Conditional independence means that, after class membership is accounted for,
the covariances/correlations between indicators are assumed to be zero.
This can be conceived of as a restricted mixture model with covariances fixed to zero.
In some cases, such constraints will be inappropriate;
for instance when the cohesion between indicators is expected to differ between classes.
As an example, a LPA of ocean plastic particles
found two classes of particles based on length and width:
a class of smaller particles with a high correlation between length and width,
meaning that these particles were approximately round or square in shape,
and a class of larger particles with a low correlation between length and width,
meaning that these particles were heterogenous in shape.
From a theoretical perspective, this makes sense,
because the smaller particles have been ground down to a more uniform shape by the elements.

It is also possible to estimate a mixture model based on latent indicators.
This means that, within each class,
one or more continuous latent variables are estimated based on the observed indicators.
Categorical latent variable membership is then estimated
based on these continuous latent variables.
A common application of this approach is in longitudinal research,
where the indicators reflect one construct assessed at different time points.
Examples of this approach include \emph{growth mixture models} (GMM)
and \emph{latent class growth analyses} (LCGA).
These techniques estimate a latent growth model
to describe individual trajectories over time.
The growth mixture model is a latent class model
where the parameters that indicate class membership
are the intercepts and variances,
and typically covariances of the latent growth variables,
e.g., a latent intercept and slope.
This technique assumes that individuals within a class can have heterogenous trajectories.
If the variance of the growth parameters is fixed to zero,
it is known as a latent class growth analysis.
This latter approach assumes that all individuals within a class
share the same identical trajectory,
and that any variance in the indicators not explained by the class-specific latent trajectories
is due to residual error variance.

The term latent class analysis originally referred to cases
where the observed indicators were categorical.
Nowadays, it is more commonly used as an umbrella term.
To prevent ambiguity, the special case
where indicators are of binary or ordinal measurement level
might be described as \emph{latent class analysis with ordinal indicators}.
Latent class models with ordinal indicators are parameterized differently from mixture models.
One common parameterization assumes that
each categorical variable reflects an underlying standard normal distribution.
The parameters are ``thresholds'' that correspond to quantiles of a standard normal distribution
(with \(N(\mu = 0, \sigma = 1)\)).
These thresholds are estimated based on
the proportion of individuals in each of the response categories of the indicator variable.
For example, a binary indicator has a single threshold that distinguishes the two response categories.
If responses are distributed 50/50, then the corresponding threshold would be \(t_1 = 0.00\).
If the responses are distributed 60/40, then the resulting threshold would be \(t_1 = 0.25\).

This paper will primarily focus on mixture models and special cases thereof,
although most of the suggested guidelines are applicable to all LCA methods.

\hypertarget{use-cases-for-latent-class-analysis}{%
\subsection{Use Cases for Latent Class Analysis}\label{use-cases-for-latent-class-analysis}}

There are several use cases for which LCA methods are suitable.
One example is to test a theory that postulates the existence of a categorical latent variable.
For example, \emph{identity status theory} posits that, at any given point in time,
adolescents reside in one of four identity statuses.
Latent class analysis can be used to identify these four statuses
based on observed indicators (e.g., self-reported identity exploration and commitment).
If results indicate that the data are better described by a different number of classes,
or that the four-class solution does not correspond to the predicted pattern of responses on the indicators,
then the theory may be called into question.

Another use case is unsupervised learning:
when the goal is to restore unobserved class membership based on observed indicators,
or to classify individuals.
For example, a mixture model can be used as a diagnostic aid
when several clinical indicators can be used to distinguish between
a fixed number of physical (Baughman, Bisgard, Lynn, \& Meade, 2006) or mental (Wu, Woody, Yang, Pan, \& Blazer, 2011) health problems.
The example of shoe size is a rudimentary illustration of this type of application.

LCA can be used as a descriptive analysis
where a researcher wishes to describe their sample
and identify a few prototypes based on many variables.
As an example, if a survey among all Dutch academics was carried out
with the goal of bringing about a funding reform,
LCA could be used to discover the types of publications that get funding.

With LCA, our goal could be to inductively identify the number of classes.
For instance, if we believe that a variable represents a group,
but don't know how many groups there are,
LCA may be an appropriate technique to answer this question.
For example, Hopfer, Tan, and Wylie (2014) studied substance use, sexual behavior,
and mental health status of urban population in Winnipeg, Canada.
The underlying assumption was that there were different risk profiles,
but their number was not known.
From a collection of indicators, LCA provided evidence
that there are four distinct risk profiles in the Winnipeg area.

Another application of LCA is to classify individuals.
In a peer harassment study,
Giang and Graham (2008) used LCA to classify over 2,000 sixth grade students
into aggressor and victim latent classes.
The five-class solution comprised of classes of
victims, aggressors, and socially adjusted students.
For instance, it revealed that there were two types of victims:
highly-victimized aggressive-victims and highly-aggressive aggressive-victims.

LCA is also appropriate when we wish to identify indicators that capture classes well.
High quality indicators will be strongly related to the latent variable
and will lead to good class separation.
This relationship of high quality indicators to the latent variable
is reflected in very high or very low conditional response probabilities.
For a simulation study exploring the effects of indicator quality on LCA, see Geiser and Wurpts (2014).
Therefore, one could use conditional response probabilities for each item to assess its quality
with regards to how well it helps separate the latent classes.
From this, a theory about the selection of indicators could be informed.

An extension of LCA is that containing covariates which can be used to predict class membership.
In this approach, we not only model the latent class variable based on indicators,
but we also relate the class membership to other explanatory variables (Vermunt (2017)).
An example of using covariates comes from Nozadi et al. (2016)
who applied LCA to identify the probability of children's membership to an anxiety class.
The authors tested several covariates including children's age, sex, and accuracy scores.
Age and sex were not found to be related to the children's latent class membership,
hence these covariates were excluded from the analysis.
Accuracy scores, however, were related to probabilities of
being in anxiety and attention-anxiety classes.
Therefore this covariate was kept as a valuable predictor.
